---
title: "OrnAtlas: Expression Atlas for Ornamental Plants"
author: "Lashika Meena"
date: "`r Sys.Date()`"
output:
  BiocStyle::html_document:
    toc: true
    toc_depth: 3
    number_sections: true
    fig_width: 8
    fig_height: 6
vignette: >
  %\VignetteIndexEntry{OrnAtlas: Expression Atlas for Ornamental Plants}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = TRUE,
  warning = FALSE,
  fig.align = "center"
)
library(OrnAtlas)
```

# Introduction

**OrnAtlas** is the first R/Bioconductor package providing an integrated
multi-study expression atlas and analysis toolkit specifically designed for
ornamental and floricultural plant species.

Key features:

- **Universal import** — works with any plant species via `importCustomAtlas()`
- **Built-in Rosa chinensis atlas** — 270 samples from 10+ BioProjects
- **Built-in annotation** — 51,302 Rosa chinensis gene annotations
- **Complete analysis pipeline** — normalize, visualize, DE analysis
- **Publication-quality figures** — botanical color themes

Currently supported species:

| Species | Common name | Genome | Samples |
|---------|------------|--------|---------|
| *Rosa chinensis* | Rose | RchiOBHm-V2 | 270 |
| *Tagetes erecta* | Marigold | ASM3086718v1 | 24 |
| Any plant species | — | User-provided | User data |

# Installation

```{r install, eval=FALSE}
if (!requireNamespace("BiocManager", quietly = TRUE))
    install.packages("BiocManager")
BiocManager::install("OrnAtlas")
```

# Quick Start — Rosa chinensis

## Import data

The `importCustomAtlas()` function is the universal entry point.
It accepts STAR `ReadsPerGene.out.tab` files from any plant species:

```{r import, eval=FALSE}
library(OrnAtlas)

# Import Rosa chinensis expression data
se <- importCustomAtlas(
  counts_dir    = "path/to/star/counts/",
  metadata_file = "path/to/SraRunTable.txt",
  species       = "Rosa chinensis",
  strandedness  = "reverse",
  min_samples   = 10
)
se
```

For demonstration, we use the built-in example dataset:

```{r example-data}
# Load built-in example data
example_files <- system.file("extdata", package = "OrnAtlas")
cat("OrnAtlas installed at:", example_files, "\n")

# Show built-in annotation
ann <- getRosaAnnotation()
cat("Rosa chinensis genes annotated:", nrow(ann), "\n")
head(ann)
```

## Normalize expression data

```{r normalize, eval=FALSE}
# CPM normalization (fast, good for visualization)
se_norm <- normalizeCounts(se, method = "CPM")

# DESeq2 VST normalization (best for DE analysis)
se_vst <- normalizeCounts(se, method = "DESeq2_vst",
                           design = ~ tissue_simple)
```

## PCA visualization

```{r pca, eval=FALSE}
# PCA of all 270 samples colored by tissue type
p <- plotPCA(se_norm,
             assay_name   = "CPM",
             color_by     = "tissue_simple",
             label_points = FALSE)
print(p)
```

The PCA reveals distinct tissue-specific transcriptional programs:

- **PC1 (38.1%)** — captures petal developmental variation
- **PC2 (12.3%)** — separates floral organ types
- Six tissue types cluster distinctly: petal, abscission zone,
  bud, meristem, stamen, pistil

## Build expression atlas

```{r atlas, eval=FALSE}
# Aggregate replicates into condition-level atlas
atlas <- buildAtlas(se_norm,
                    group_var  = "tissue_simple",
                    assay_name = "CPM")

# Heatmap of top 50 variable genes
plotAtlas(atlas,
          top_n  = 50,
          title  = "Rosa chinensis Expression Atlas")
```

## Differential expression analysis

```{r de, eval=FALSE}
# DE analysis: petal vs abscission zone
results <- runDE(
  se,
  design        = ~ tissue_simple,
  contrast      = c("tissue_simple", "petal", "abscission zone"),
  alpha         = 0.05,
  lfc_threshold = 1
)

# Summary
cat("Upregulated in petal:", sum(results$direction == "Up"), "\n")
cat("Downregulated in petal:", sum(results$direction == "Down"), "\n")

# Volcano plot
plotVolcano(results,
            title = "Petal vs Abscission Zone - Rosa chinensis",
            top_n = 20)
```

## Annotate DE results

```{r annotate, eval=FALSE}
# Add official locus names to DE results
results_ann <- annotateDEResults(results)

# View top significant genes with annotation
top_genes <- results_ann[results_ann$significant == TRUE, ]
top_genes <- top_genes[order(top_genes$padj), ]
head(top_genes[, c("gene_id", "locus_name",
                    "log2FoldChange", "padj",
                    "direction")], 10)
```

# Universal Import — Any Plant Species

The key innovation of OrnAtlas is `importCustomAtlas()` which works
for **any plant species** with STAR-aligned RNA-seq data.

## Example: Tagetes erecta (Marigold)

Marigold is India's most economically important floriculture crop,
widely used in religious ceremonies, festivals, and as a source of
lutein for pharmaceutical production.

```{r marigold, eval=FALSE}
# Import marigold expression data
se_marigold <- importCustomAtlas(
  counts_dir    = "path/to/marigold/counts/",
  metadata_file = "path/to/marigold_metadata.csv",
  species       = "Tagetes erecta",
  sample_col    = "Run",
  strandedness  = "reverse",
  min_samples   = 3
)

# Normalize
se_mg_norm <- normalizeCounts(se_marigold, method = "CPM")

# PCA colored by developmental stage and cultivar
p_mg <- plotPCA(se_mg_norm,
                assay_name   = "CPM",
                color_by     = "dev_stage",
                shape_by     = "flower_color",
                label_points = FALSE)
print(p_mg)

# DE: orange (V-01) vs yellow mutant (V-01M) at Stage IV
results_mg <- runDE(
  se_marigold,
  design        = ~ flower_color,
  contrast      = c("flower_color", "orange", "yellow"),
  alpha         = 0.05,
  lfc_threshold = 1
)

plotVolcano(results_mg,
            title = "Orange vs Yellow - Tagetes erecta Stage IV",
            top_n = 15)
```

## Required input format

OrnAtlas accepts output from the STAR aligner with
`--quantMode GeneCounts`. Each sample generates a
`ReadsPerGene.out.tab` file:

```
N_unmapped      1234    1234    1234
N_multimapping  5678    5678    5678
N_noFeature     9012    9012    9012
N_ambiguous     3456    3456    3456
gene1           0       0       25
gene2           15      8       7
gene3           234     118     116
```

The metadata CSV must have a column of sample IDs matching the
count file names, plus any experimental variables:

```
Run,species,tissue,dev_stage,cultivar,replicate
SRR001,Rosa chinensis,petal,Stage_I,Old_Blush,1
SRR002,Rosa chinensis,petal,Stage_I,Old_Blush,2
SRR003,Rosa chinensis,leaf,mature,Old_Blush,1
```

# Rosa chinensis Gene Annotation

OrnAtlas includes a built-in annotation database for all 51,302
genes in the RchiOBHm-V2 reference genome:

```{r annotation}
# Get full annotation
ann <- getRosaAnnotation()
cat("Total genes:", nrow(ann), "\n")
table(ann$gene_biotype)
```

```{r annotate-genes}
# Annotate a list of gene IDs
my_genes <- c("gene24707", "gene47555", "gene4853",
               "gene28236", "gene29625")
annotateGenes(my_genes)
```

# Session Information

```{r session}
sessionInfo()
```
